Efficient Computation of Extended Sources

### 2021-05-29 repro/papers/sseprecond/paper.tex
interim

The preconditioner construction is very similar to the time-reversal
method in photoacoustic tomography
\cite[]{StefanovUhlmannIP:09}. Preconditioning amounts to a change of
norm in the domain and range spaces of the modeling operator. In this
case, the modfied norms are weighted $L^2$, and the weight operators
map pressure to corresponding surface source on the source and
receiver planes. These are essentially the same as the ``hyperbolic
Dirichlet-to-Neumann'' map that plays a prominent role in
thermoacoustic tomography and other wave inverse problems
\cite[]{Rachele:00,StefUhl:05}. \cite{HouSymes:EAGE16} demonstrated a
very similar preconditioner for Least Squares Migration, also for its
subsurface offset extension \cite[]{HouSymes:16}, motivated by
\cite{tenKroode:12}. These constructions also involve the
Dirichlet-to-Neumann operator. This concept also turns up in hidden
form in the work of Yu Zhang and collaborators on true amplitude
migration
\cite[]{YuZhang:14,TangXuZhang:13,XuWang:2012,XuZhangTang:11,Zhang:SEG09,Zhang
  YuSun:08,ZhangSunGray:07,ZhangBleistein:05,Bleisteinetal:05}.

@@

The negative jump $-[v^{\pm}_{z}]|_{z=z_s}$ is thus a function of $\phi$. Define
the {\em pressure-to-source} operator $\Lambda^{\pm}_{z_s}$ by
\begin{equation}
  \label{eqn:deflam}
  \Lambda^{\pm}_{z_s}\phi = -[v^{\pm}_{1,z}]|_{z=z_s}
\end{equation}
The conclusion: if $h_s = \Lambda^{\pm}_{z_s}\phi$ and $f_s=0$ in the
system \ref{eqn:awepm}, then $\phi=P_sp^{\pm}$.

@@

The analysis also reveals that the solution components not continuous
at $z=z_s$ are odd there:
\begin{equation}
  \label{eqn:odd1}
  \lim_{z\rightarrow z_s^+} v^{\pm}_{z} \approx - \lim_{z\rightarrow z_s^-}
  v^{\pm}_{z}
\end{equation}
for the solution of \ref{eqn:awepm} with $f_s=0$.
Similarly, 
\begin{equation}
  \label{eqn:odd2}
  \lim_{z\rightarrow z_s^+} p^{\pm}\approx - \lim_{z\rightarrow z_s^-}
  p^{\pm}
\end{equation}
for the solution of \ref{eqn:awepm} with $h_s=0$. Here ``$\approx$''
means in the sense of high frequency asymptotics, that is, that the
difference between the two sides is relatively smooth, hence small if
the data is highly oscillatory. Therefore if $f_s=0$ in system \ref{eqn:awepm},
\begin{equation}
  h_s = \Lambda^{\pm}_{z_s}P_sp^{\pm} = -[v^{\pm}_{z}]|_{z=z_s} \approx -2
  P_sv^{\pm}_{z}
  \label{eqn:tracejump10}
\end{equation}
Similarly, if $h_s=0$ in system \ref{eqn:awepm}, then
\begin{equation}
  \label{eqn:tracejump20}
  f_s = -[p^{\pm}]|_{z=z_s} \approx -2 P_s p^{\pm}.
\end{equation}
Thus $f_s$ determines approximately the boundary value of $p^{\pm}$,
as a solution of the acoustic wave system in the half-space
$z>z_s$. However, as repeated in equation \ref{eqn:tracejump10}, a
solution with this boundary value is also the restriction to $z>z_s$
of a solution to \ref{eqn:awepm} with $f_s=0$ and $h_s=
\Lambda^{\pm}_{z_s}P_sp^{\pm}$. Therefore if
\begin{equation}
  \label{eqn:hfcondn}
  h_s =\Lambda^{\pm}_{z_s}f_s,
\end{equation}
then the pressure boundary value $P_sp^{\pm}$ is the
same for the solutions of \ref{eqn:awepm} for source vectors $(h_s,0)$
and $(0,f_s)$. Since the pressure boundary values are the same, the solutions
in $z>z_s$ are the same. In particular, since $z_r>z_s$ and ${\cal
  S}^{\pm}_{z_s,z_r}(h_s,f_s)^T = (P_rp^{\pm},P_rv^{\pm}_z)^T$, it follows
that
\begin{equation}
  \label{eqn:snull}
  {\cal S}^{\pm}_{z_s,z_r}(-\Lambda^{\pm}_{z_s}f_s,f_s)^T \approx 0.
\end{equation}

@@

For a point source at $z_d=3500$ m, $x_d=3500$ m, same bandpass filter
wavelet as used for prior examples, causal data on $z=z_s=3000$ m are
shown in Figures \ref{fig:dsrcphh0} and \ref{fig:dsrcvzhh0}. Since the
mechanical parameters in the homogeneous and lens models are the same
for $z<z_s$, and no rays return to this zone in either model, these
data are asymptotically the same for both models, and I show only the
homogenous medium results. Regard these gathers as the trace
$(P_sp^+,P_sv^+_z)$ on $z=z_s$ of a downgoing acoustic field in $z>z_s$.
Equations \ref{eqn:tracejump10} and
\ref{eqn:tracejump20} show that these differ by a factor of -2 from
source functions $f_s$ and $h_s$ as in the system \ref{eqn:awepm},
with $h_s=0$ and $f_s=0$ respectively.

@@

\plot{dsrcpghh0}{width=\textwidth}{Trace $P_sp^+$ on $z=z_s=3000$ m of
  pressure field from point source at $z_d=3500$ m, $x_d=3500$ m,
  bandpass filter source.}

\plot{dsrcvzghh0}{width=\textwidth}{Trace $P_sv_z^+$ on $z=z_s=3000$ m of
  vertical velocity field from point source at $z_d=3500$ m, $x_d=3500$ m,
  bandpass filter source.}

\plot{dfwdphh0}{width=\textwidth}{Pressure gather at receiver depth 
  $z_r=1000$ m from field generated by causal solution of acoustic 
  system \ref{eqn:awepm} in the homogeneous model described in the 
  text, with extended pressure source (constitutive defect) on $z=z_s=3000$ m given by $-2$
  times the field depicted in Figure \ref{fig:dsrcvzhh0}
  ($h_s=-2P_sv_z^+$) and zero velocity source (vertical load) 
  ($f_s=0$).}

\plot{daltphh0}{width=\textwidth}{Pressure gather at receiver depth 
  $z_r=1000$ m from field generated by causal solution of acoustic 
  system \ref{eqn:awepm} in the homogeneous model described in the 
  text, with extended velocity source (vertical load) on $z=z_s=3000$ m given by $-2$
  times the field depicted in Figure \ref{fig:dsrcphh0}
  ($f_s=-2P_sp^+$) and zero pressure source (constitutive defect)
  ($h_s=0$).}

\plot{dfwdplh0}{width=\textwidth}{Pressure gather at receiver depth 
  $z_r=1000$ m from field generated by causal solution of acoustic 
  system \ref{eqn:awepm} in the lens model described in the 
  text, with extended pressure source (constitutive defect) on $z=z_s=3000$ m given by $-2$
  times the field depicted in Figure \ref{fig:dsrcvzhh0}
  ($h_s=-2P_sv_z^+$) and zero velocity source (vertical load) 
  ($f_s=0$).}

\plot{daltplh0}{width=\textwidth}{Pressure gather at receiver depth 
  $z_r=1000$ m from field generated by causal solution of acoustic 
  system \ref{eqn:awepm} in the lens model described in the 
  text, with extended velocity source (vertical load) on $z=z_s=3000$ m given by $-2$
  times the field depicted in Figure \ref{fig:dsrcphh0}
  ($f_s=-2P_sp^+$) and zero pressure source (constitutive defect)
  ($h_s=0$).}

@@

Computation of $\Lambda^{\pm}_{z_s}$ will prove a critical step in an
efficient iterative solution of the linear subproblem
\ref{eqn:esis}. Direct computation, for instance by solving
\ref{eqn:sawe2} and reading off $P_sv^{\pm}_{2,z}$, turns out to be
numerically ill-behaved. The relation \ref{eqn:snull} provides and
alternative approach, {\em provided} that an accurate approximate
inverse to $S^+_{z_s,z_r}$ is available. For instance, the first row
of the vector equation \ref{eqn:snull} reads (slightly rearranged):
\[
\Pi_0{\cal S}^+_{z_s,z_r}\Pi_1 f_s \approx \Pi_0{\cal
  S}^+_{z_s,z_r}Pi_0\Lambda^+_{z_s}f_s
\]
\begin{equation}
  \label{eqn:lamidea}
  = S^++_{z_s,z_r}\Lambda^+_{z_s}f_s.
\end{equation}
An approximate inverse for $S^++_{z_s,z_r}$ would permit (approximate)
solution of this equation for $\Lambda^+_{z_s}f_s$.

Recall that the source vector $(h_s,f_s)$ is assumed to produce a
downgoing field $(p^+,\bv^+)$, that is, emanates high-frequency energy only along
rays that make an angle with the vertical bounded below by a common
minimum angle. Such rays leave $\Omega$ within a common maximum time.

An immediate consequence of the downgoing assumption stated above is that in the
slab $z_s<z<z_r$, the field $(p^+,\bv^+)$ approximates the solution of an
anti-causal evolution equation. Indeed, for any choice of $T$, define
$(\tilde{p}^-,\tilde{\bv}^-)$ to be the solution in the half-space
$\Omega \times (-\infty,T]$ of
\begin{eqnarray}
\label{eqn:revawe}
  \frac{1}{\kappa}\frac{\partial \tilde{p}^-}{\partial t} & = & - \nabla \cdot \tilde{\bv}^-, \nonumber \\
  \rho\frac{\partial \tilde{\bv}^-}{\partial t} & = & - \nabla \tilde{p}^-,\nonumber \\
  \tilde{p}^- & =& 0,  \mbox{ for } t=T,\nonumber\\ 
  \tilde{\bv}^- & = & 0 \mbox{ for } t=T,\nonumber\\
  P_r\tilde{p}^- &=& P_rp^+ \mbox { for } t \le T . 
\end{eqnarray}
That is, $\tilde{p}^-$ has the same boundary value as
$p^+$ for $t<T$, but $(\tilde{p}^-,\tilde{\bv}^-)$ vanishes at
$t=T$. Suppose that $T$ is chosen so that the rays carrying
significant energy in $(p^+,\bv^+)$ have left the half-space $\Omega$
at time $T$. Then

As will be shown later, under the downgoing assumption, $P_r p^+
\approx P_r \tilde{p}^+$.
Therefore $(\delta p,\delta \bv)=(\tilde{p}^+-\tilde{p}^-,\tilde{\bv}^+-\tilde{\bv}^-)$
differs negligibly from the solution of
\begin{eqnarray}
\label{eqn:deltaawe}
  \frac{1}{\kappa}\frac{\partial \delta p}{\partial t} & = & -
                                                               \nabla
                                                               \cdot
                                                               \delta \bv, \nonumber \\
  \rho\frac{\partial \delta \bv}{\partial t} & = & - \nabla \delta p,\nonumber \\
  \delta p & =& \tilde{p}^+,  \mbox{ for } t=T,\nonumber\\ 
  \delta \bv & = & \tilde{\bv}^+ \mbox{ for } t=T,\nonumber\\
  P_r\delta p & = & 0,
\end{eqnarray}
in $\Omega \times \bR$. Since
$E_{\Omega}[(\tilde{p}^+,\tilde{\bv}^+)](T) \rightarrow 0$ as $T
\rightarrow \infty$, the usual energy estimate implies that the energy $\delta
E$ in this field,
\[
  \delta E_{\Omega}(t) = \frac{1}{2} \int_{\Omega} \,d\bx \,
 \left(\frac{(\delta p)^2}{\kappa} + \rho |\delta \bv|^2\right)
\]
satisfies $\delta E_{\Omega}(t) \rightarrow 0$ as $T \rightarrow
\infty$. This decay does not generally imply that $P_s\delta p, P_s \delta \bv)
\rightarrow 0$ (for instance) as $T \rightarrow \infty$, since the trace operator
$P_s$ is not continuous, but it does so imply for purely downgoing
fields, as will be demonstrated later, so
\begin{equation}
  \label{eqn:approxinv0}
  \lim_{T \rightarrow \infty} P_s\tilde{v}_z^- \approx \lim_{z \rightarrow
    0^+}\tilde{v}_z^+ \approx \lim_{z \rightarrow 0^+}v^+_z
\end{equation}

Recall (equations \ref{eqn:deflam}, \ref{eqn:jump0}) that
\[
  \Lambda^+_{z_s} P_s p^+ \approx 2 \lim_{z \rightarrow z_s^+}v^+_z 
\]
and that with $h_s= \Lambda^+_{z_s}P_sp^+$,
\[
  S^+_{z_s,z_r}h_s = P_r p^+
\]
Together with the last two equations, the relation
\ref{eqn:approxinv0} implies that
\begin{equation}
  \label{eqn:approxinv1}
  S^+_{z_s,z_r}(2 P_s \tilde{v}_z^-) \approx P_rp^+.
\end{equation}
As shown in the last section, the solution of the system
\ref{eqn:revawe} differs negligibly from the anti-causal solution $(p^-_2,\bv^-_2)$
of the system \ref{eqn:sawe2} with (i) $z_s$ replaced by $z_r$, and
(ii) $\psi = [p^-_2]|_{z=z_r} = -2 \lim_{z \rightarrow z_r^-} p^-_2
\approx -2 P_rp^+$. Therefore (definitions \ref{eqn:sdef}, \ref{eqn:vdef})
\begin{equation}
  \label{eqn:approxinv2}
  P_s \tilde{v}_z^- \approx V^-_{z_r,z_s}(-2 P_rp^+).
\end{equation}
Taken together, relations \ref{eqn:approxinv1} and
\ref{eqn:approxinv2} yield the first equation in the first main result of this section:
\begin{eqnarray}
  \label{eqn:approxinv}
  V^-_{z_r,z_s} S^+_{z_s,z_r} & \approx & -\frac{1}{4}I, \nonumber\\
  S^-_{z_r,z_s} V^+_{z_s,z_r} & \approx & -\frac{1}{4}I, \nonumber\\
  V^+_{z_s,z_r} S^-_{z_r,z_s} & \approx & -\frac{1}{4}I, \nonumber\\
  S^+_{z_s,z_r} V^-_{z_r,z_s} & \approx & -\frac{1}{4}I.\\
\end{eqnarray}.
The second equation follows by an exactly analogous argument, and the
last two by time reversal and interchange of the roles of $z_s$ and
$z_r$.

Having identified approximate two-sided inverses for $S^{\pm},
V^{\pm}$, the approximate null space identity \ref{eqn:lamnull} leads
to a relation between $\Lambda^{\pm}$ and ${\cal S}^{\pm}$ that
provides the foundation for practical computations. Applying $\Pi_0$
to both sides of \ref{eqn:lamnull} gives
\[
 \Pi_0 {\cal S}^+_{z_s,z_r}\Pi_1^T \approx \Pi_0 {\cal S}^+_{z_s,z_r}\Pi_0^T
 \Lambda^+_{z_s} = S^+_{z_s,z_r} \Lambda^+_{z_s}.
\]
From the first approximate identity in \ref{eqn:approxinv},
\begin{equation}
  \label{eqn:lamident}
  V^{-}_{z_r,z_s}\Pi_0 {\cal S}^+_{z_s,z_r}\Pi_1^T \approx -\frac{1}{4}\Lambda^+_{z_s}.
\end{equation}
This identity is the second major result of this section: it shows how
to compute that action of $\Lambda^+_{z_s}$ by propagating the input
pressure trace, identified as a source for the velocity evolution,
forward in time from $z_s$ to $z_r$
reading off the pressure trace on $z=z_r$, identifying it once more as
a source for velocity, propagating it backwards in time from $z_r$ to
$z_s$, and finally reading off the velocity trace, interpreted as a
pressure evolution source on $z_s$. I will give a detailed description
of this computation in the section on numerical results.

@@

Computation of the transpose of $\Lambda^+$ (exact, not approximate in
the high frequency sense) will be important in the
construction of a preconditioner. The relation \ref{eqn:lamident} also
provides an computation for this operator. The idea is to treat the
left-hand side of equation \ref{eqn:lamident} as an approximation to
$\Lambda$, and compute its exact transposes via the adjoint state
method (equations \ref{eqn:trtr} \ref{eqn:trtrcomp}):
\begin{equation}
  \label{eqn:lamtransp}
   \Pi_1 {\cal S}^-_{z_r,z_s}\Pi_0^T V^{+}_{z_s,z_r} \approx -\frac{1}{4}(\Lambda^+_{z_s})^T.
\end{equation} 
The left-hand sides of \ref{eqn:lamident} and \ref{eqn:lamtransp} will
simply be substituted for $\Lambda$ and its transpose in the
preconditioner construction: preconditioners must be (precisely)
self-adjoint, but need only be approximated to perform well.

@@

>>>>

The preconditioner construction is very similar to the time-reversal
method in photoacoustic tomography
\cite[]{StefanovUhlmannIP:09}. Preconditioning amounts to a change of
norm in the domain and range spaces of the modeling operator. In this
case, the modfied norms are weighted $L^2$, and the weight operators
map pressure to corresponding surface source on the source and
receiver planes. These are essentially the same as the ``hyperbolic
Dirichlet-to-Neumann'' map that plays a prominent role in
thermoacoustic tomography and other wave inverse problems
\cite[]{Rachele:00,StefUhl:05}. \cite{HouSymes:EAGE16} demonstrated a
very similar preconditioner for Least Squares Migration, also for its
subsurface offset extension \cite[]{HouSymes:16}, motivated by
\cite{tenKroode:12}. These constructions also involve the
Dirichlet-to-Neumann operator. This concept also turns up in hidden
form in the work of Yu Zhang and collaborators on true amplitude
migration
\cite[]{YuZhang:14,TangXuZhang:13,XuWang:2012,XuZhangTang:11,Zhang:SEG09}.

@@

The negative jump $-[v^{\pm}_{z}]|_{z=z_s}$ is thus a function of $\phi$. Define
the {\em pressure-to-source} operator $\Lambda^{\pm}_{z_s}$ by
\begin{equation}
  \label{eqn:deflam}
  \Lambda^{\pm}_{z_s}\phi = -[v^{\pm}_{z}]|_{z=z_s}
\end{equation}
The conclusion: if $h_s = \Lambda^{\pm}_{z_s}\phi$ and $f_s=0$ in the
system \ref{eqn:awepm}, then $\phi=P_sp^{\pm}$.

@@

The analysis also reveals that the solution components not continuous
at $z=z_s$ are odd there:
\begin{equation}
  \label{eqn:odd1}
  \lim_{z\rightarrow z_s^+} v^{\pm}_{z} \approx - \lim_{z\rightarrow z_s^-}
  v^{\pm}_{z}
\end{equation}
for the solution of \ref{eqn:awepm} with $f_s=0$.
Similarly, 
\begin{equation}
  \label{eqn:odd2}
  \lim_{z\rightarrow z_s^+} p^{\pm}\approx - \lim_{z\rightarrow z_s^-}
  p^{\pm}
\end{equation}
for the solution of \ref{eqn:awepm} with $h_s=0$. Here ``$\approx$''
means in the sense of high frequency asymptotics, that is, that the
difference between the two sides is relatively smooth, hence small if
the data is highly oscillatory. Therefore if $f_s=0$ in system \ref{eqn:awepm},
\begin{equation}
  h_s = \Lambda^{\pm}_{z_s}P_sp^{\pm} = -[v^{\pm}_{z}]|_{z=z_s} \approx -2
  P_sv^{\pm}_{z}
  \label{eqn:tracejump10}
\end{equation}
Similarly, if $h_s=0$ in system \ref{eqn:awepm}, then
\begin{equation}
  \label{eqn:tracejump20}
  f_s = -[p^{\pm}]|_{z=z_s} \approx -2 P_s p^{\pm}.
\end{equation}
Thus $f_s$ determines approximately the boundary value of $p^{\pm}$,
as a solution of the acoustic wave system in the half-space
$z>z_s$. However, as repeated in equation \ref{eqn:tracejump10}, a
solution with this boundary value is also the restriction to $z>z_s$
of a solution to \ref{eqn:awepm} with $f_s=0$ and $h_s=
\Lambda^{\pm}_{z_s}P_sp^{\pm}$. Therefore if
\begin{equation}
  \label{eqn:hfcondn}
  h_s =-\frac{1}{2}\Lambda^{\pm}_{z_s}f_s,
\end{equation}
then the pressure boundary value $P_sp^{\pm}$ is the
same for the solutions of \ref{eqn:awepm} for source vectors $(h_s,0)$
and $(0,f_s)$. Since the pressure boundary values are the same, the solutions
in $z>z_s$ are the same. In particular, since $z_r>z_s$ and ${\cal
  S}^{\pm}_{z_s,z_r}(h_s,f_s)^T = (P_rp^{\pm},P_rv^{\pm}_z)^T$, it follows
that
\begin{equation}
  \label{eqn:snull}
  {\cal S}^{\pm}_{z_s,z_r}\left(-\frac{1}{2}\Lambda^{\pm}_{z_s}f_s,f_s\right)^T \approx 0.
\end{equation}

@@

Since the
mechanical parameters in the homogeneous and lens models are the same
for $z<z_s$, and no rays return to this zone in either model, these
data are asymptotically the same for both models, and I show only the
homogenous medium results. Regard these gathers as the trace
$(P_sp^+,P_sv^+_z)$ on $z=z_s$ of a downgoing acoustic field in $z>z_s$.
Equations \ref{eqn:tracejump10} and
\ref{eqn:tracejump20} show that these differ by a factor of -2 from
source functions $f_s$ and $h_s$ as in the system \ref{eqn:awepm},
with $h_s=0$ and $f_s=0$ respectively. For a point source at $z_d=3500$ m, $x_d=3500$ m, same bandpass filter
wavelet as used for prior examples, causal data on $z=z_s=3000$ m are
shown in Figures \ref{fig:dsrcpghh0} and \ref{fig:dsrcvzghh0}.

@@

@@

Computation of $\Lambda^{\pm}_{z_s}$ will prove a critical step in an
efficient iterative solution of the linear subproblem
\ref{eqn:esis}. Direct computation, for instance by solving
\ref{eqn:sawe2} and reading off $P_sv^{\pm}_{2,z}$, turns out to be
numerically ill-behaved. The relation \ref{eqn:snull} provides and
alternative approach, {\em provided} that an accurate approximate
inverse to $S^+_{z_s,z_r}$ is available. For instance, the first row
of the vector equation \ref{eqn:snull} reads (slightly rearranged):
\[
\Pi_0{\cal S}^+_{z_s,z_r}\Pi_1^T f_s \approx \frac{1}{2}\Pi_0{\cal
  S}^+_{z_s,z_r}\Pi_0\Lambda^+_{z_s}f_s
\]
\begin{equation}
  \label{eqn:lamidea}
  = \frac{1}{2} S^+_{z_s,z_r}\Lambda^+_{z_s}f_s.
\end{equation}
An approximate inverse for $S^++_{z_s,z_r}$ would permit (approximate)
solution of this equation for $\Lambda^+_{z_s}f_s$.

Recall that the source vector $(h_s,f_s)$ is assumed to produce a
downgoing field $(p^+,\bv^+)$, that is, emanates high-frequency energy only along
rays that make an angle with the vertical bounded below by a common
minimum angle. Such rays leave $\Omega$ within a common maximum
time. Consequently (Appendix B), in the
slab $z_s<z<z_r$, the field $(p^+,\bv^+)$ approximates the solution of an
anti-causal evolution equation. Choose $\chi(t)$ to be a smooth function
that is $= 0$ for $t \gg 0$ and $=1$ at times when near rays carrying
high-frequency energy in $(p^+,\bv^+)$ cross $z=z_r$. Define 
$(\tilde{p}^-,\tilde{\bv}^-)$ to be the solution in the half-space
$\Omega \times \bR$ of
\begin{eqnarray}
\label{eqn:revawe}
  \frac{1}{\kappa}\frac{\partial \tilde{p}^-}{\partial t} & = & - \nabla \cdot \tilde{\bv}^-, \nonumber \\
  \rho\frac{\partial \tilde{\bv}^-}{\partial t} & = & - \nabla \tilde{p}^-,\nonumber \\
  \tilde{p}^- & =& 0,  \mbox{ for } t \gg 0\\ 
  \tilde{\bv}^- & = & 0 \mbox{ for } t \gg 0\\
  P_r\tilde{p}^- &=& \chi P_rp^+ . 
\end{eqnarray}
That is, $\tilde{p}^-$ has the same boundary value on $z=z_r$ as
$p^+$, except for low-frequency residue that is muted by
$\chi$. Therefore
$p^+ \approx \tilde{p}^-, \bv^+ \approx \tilde{\bv}^-$ near
$z=z_r$. Since the right-hand sides in the system \ref{eqn:awepm} are
singular only on $z=z_s$, and the high-frequency components of
$(p^+,\bv^+)$ are carried by downgoing rays, these differ negligibly
from the the high-frequency components of
$(\tilde{p}^-,\tilde{\bv}^-)$ in the space-time slab $z_s<z<z_r$, and
the approximation holds throughout this region. In particular
$P_sv^+_z \approx P_s \tilde{v}^-_z$. In view of the relation
\ref{eqn:tracejump10},
\begin{equation}
  \label{eqn:tildevtohsubs}
  -2P_s\tilde{v}^-_z \approx h_s,
\end{equation}
so solution
of the system \ref{eqn:revawe} followed by restriction to $z=z_s$ and
multiplication by $-2$ 
approximately inverts the map $S^+{z_s,z_r}: h_s \mapsto P_rp^+$.

Next observe that in view of the relation \ref{eqn:tracejump20}, and
the downgoing nature of the ray system carrying the high frequency
energy in $(p^+,\bv^+)$, the field $(\tilde{p}^-,\tilde{\bv}^-)$ is
actually the restriction to $z<z_r$ of the anti-causal solution of \ref{eqn:awepm}
with $z_s$ replaced by $z_r$, zero constitutive defect, and vertical
load given by the jump in pressure at $z=z_r$ - for this field, use
the same notation. Continuity of vertical
velocity $\tilde{v}^-_z$ at $z=z_r$ implies that the vertical load is
\[
  f_r = -[\tilde{p}^-]|_{z=z_r} = 2 P_r \tilde{p}^- = 2 P_r p^+
\]
(from the definition \ref{eqn:defsamp}, $P_r$ is the limit from the
left). Thus
\[
  P_s \tilde{v^-_z} \approx V^-_{z_r,z_s}(2 P_rp^+) \approx
  2V^-_{z_r,z_s}S^+_{z_r,z_s}h_s.
\]
so
\[
  h_s \approx -2 P_s v^+_z \approx -2 P_s \tilde{v}^-_z \approx
  -4V^-_{z_r,z_s}S^+_{z_r,z_s}h_s
\]
Combine this observation with \ref{eqn:tildevtohsubs} to obtain
\[
 -4  V^-_{z_r,z_s} S^+_{z_s,z_r}  \approx  I,
\]
This relation combines with the identity \ref{eqn:trtrcomp} to
yield the first main result of this section:
\begin{eqnarray}
  \label{eqn:approxinv}
  (V^+_{z_s,z_r})^T S^+_{z_s,z_r} & \approx & \frac{1}{4}I, \nonumber\\
  (S^+_{z_s,z_r})^T V^+_{z_s,z_r} & \approx & \frac{1}{4}I, \nonumber\\
  V^+_{z_s,z_r} (S^+_{z_s,z_r})^T & \approx & \frac{1}{4}I, \nonumber\\
  S^+_{z_s,z_r} (V^+_{z_s,z_r})^T & \approx & \frac{1}{4}I.\\
\end{eqnarray}.
The second equation is simply the transpose of the first, and the
last two follow by by an exactly analogous argument using time
reversal and interchange of the roles of $z_s$ and$z_r$.

The conclusion is significant enough to merit restating in English:
provided that high-frequency energy in the various fields is carried
along downgoing ray fields, the transpose of $V^+$ is an approximate
inverse to $S^+$, modulo a factor of 4. To recover the pressure source
$h_s$ generating a pressure gather $P_rp$ at $z=z_r$, multiply the
latter by -2, then apply the transpose of $V^+_{z_s,z_r}$ to this
gather, reading out a vertical velocity field at $z=z_s$. Multiply
again by -2 and you have a high-frequency approximation to $h_s$.

I have applied this procedure to the pressure gathers shown in Figures
\ref{fig:recphh0} and \ref{fig:recplh0}, generated by a point source
at $z_s=3000$ m, $x_s=3500$ m, propagating in the homogeneous and lens
models respectively. The source time dependence is again a
[1, 2.5, 10, 12.5] Hz trapezoidal bandpass filter. As noted earlier,
a point source at depth $z=z_s$ generates a non-downgoing acoustic
field in the slab $z_s < z < z_r$. However the pressure gather at
$z=z_r$ contains only downgoing energy after a suitable mute: this is
evident in the figures. Therefore one would expect inversion of this
data to produce an aperture-limited (in effect, dip-filtered) version
of the point source.

Figure \ref{fig:srcvzghh0} shows the approximate inversion suggested
by the first equation in \ref{eqn:approxinv}. The result indeed looks
like a dip-filtered point source. To see that this is indeed an
inversion, I apply $S^{+}_{z_s,z_r}$ to the source shown in Figure
\ref{fig:srcvzghh0}, to produce the re-simulation shown in Figure
\ref{fig:fwdphh0}. Comparison with the gather in Figure
\ref{fig:recphh0} via a difference plot \ref{fig:diffrecphh0} verifies
that a very close match has been obtained, so the source in Figure
\ref{fig:srcvzghh0} is indeed a satisfactory inversion, even though
the original point source has not been recovered: the difference is in
the null space of the modeling operator, arising from the limited
aperture of the acquisition geometry.

As mentioned at the beginning of this section, the availability of approximate two-sided inverses for $S^{+},
V^{+}$ leads to a practical computation for $\Lambda^{+}$: apply
$4(V^{+}_{z_s,z_r})^T$ to both sides of equation \ref{eqn:lamidea} and
use the first equation in the list \ref{eqn:approxinv} to get
\begin{equation}
  \label{eqn:lamident}
  \Lambda^+_{z_s} \approx 8(V^{+}_{z_s,z_r})^T\Pi_0 {\cal S}^+_{z_s,z_r}\Pi_1^T.
\end{equation}
This identity is the second major result of this section: it shows how
to compute that action of $\Lambda^+_{z_s}$ by propagating the input
pressure trace, identified as a source for the velocity evolution,
forward in time from $z_s$ to $z_r$
reading off the pressure trace on $z=z_r$, identifying it once more as
a point load (source for velocity), propagating it backwards in time from $z_r$ to
$z_s$, and finally reading off the velocity trace, interpreted as a
pressure evolution source on $z_s$.

@@

Computation of the transpose of $\Lambda^+$ (exact, not approximate in
the high frequency sense) is critical to the successful
construction of the preconditioner. The relation \ref{eqn:lamident}
does not provide a computation for this operator. However set
\begin{equation}
  \label{eqn:lamtilde}
  \tilde{\Lambda}^+_{z_s} = 8(V^{+}_{z_s,z_r})^T\Pi_0 {\cal S}^+_{z_s,z_r}\Pi_1^T.
\end{equation}
Then \ref{eqn:lamident} can be rewritten
\[
  \Lambda^+_{z_s} \approx \tilde{\Lambda}^+_{z_s}.
\]
Since successful preconditioning requires only approximate inversion,
use of $\tilde{\Lambda}^+_{z_s} $ in place of $\Lambda^+_{z_s}$ will
still yield a working preconditioner, and the former can be transposed
to machine precision via the definition \ref{eqn:lamtilde} and the adjoint state
method (equations \ref{eqn:trtr} \ref{eqn:trtrcomp}):
\begin{equation}
  \label{eqn:lamtransp}
   (\tilde{\Lambda}^+_{z_s})^T = 8 \Pi_1 {\cal S}^-_{z_r,z_s}\Pi_0^T V^{+}_{z_s,z_r}
\end{equation}

@@

\plot{dsrcpghh0}{width=\textwidth}{Trace $P_sp^+$ on $z=z_s=3000$ m of
  pressure field from point source at $z_d=3500$ m, $x_d=3500$ m,
  bandpass filter source. Scaled to represent point load (velocity
  source). }

\plot{dsrcvzghh0}{width=\textwidth}{$\Lambda^+_{z_s}P_sp^+$ = scaled
  version of trace $P_sv_z^+$ on $z=z_s=3000$ m of
  vertical velocity field from point source at $z_d=3500$ m, $x_d=3500$ m,
  bandpass filter source.}

\plot{dfwdphh0}{width=\textwidth}{Pressure gather at receiver depth 
  $z_r=1000$ m from field generated by causal solution of acoustic 
  system \ref{eqn:awepm} in the homogeneous model described in the 
  text, with extended pressure source (constitutive defect) $h_s$ on $z=z_s=3000$ m given by  field depicted in Figure \ref{fig:dsrcvzghh0}
  ($h_s=-2P_sv_z^+=\Lambda^+_{z_s}P_sp^+$) and zero velocity source (vertical load) 
  ($f_s=0$).}

\plot{daltphh0}{width=\textwidth}{Pressure gather at receiver depth 
  $z_r=1000$ m from field generated by causal solution of acoustic 
  system \ref{eqn:awepm} in the homogeneous model described in the 
  text, with extended velocity source (vertical load) $f_s$ on $z=z_s=3000$
  m given by the field depicted in Figure \ref{fig:dsrcpghh0}
  ($f_s=-2P_sp^+$) and zero pressure source (constitutive defect)
  ($h_s=0$).}

\plot{dfwdplh0}{width=\textwidth}{Pressure gather at receiver depth 
  $z_r=1000$ m from field generated by causal solution of acoustic 
  system \ref{eqn:awepm} in the lens model described in the 
  text, with extended pressure source (constitutive defect) on $z=z_s=3000$ m given bythe field depicted in Figure \ref{fig:dsrcvzghh0}
   ($h_s=-2P_sv_z^+=\Lambda^+_{z_s}P_sp^+$) and zero velocity source (vertical load) 
  ($f_s=0$).}

\plot{daltplh0}{width=\textwidth}{Pressure gather at receiver depth 
  $z_r=1000$ m from field generated by causal solution of acoustic 
  system \ref{eqn:awepm} in the lens model described in the 
  text, with extended velocity source (vertical load) on $z=z_s=3000$
  m given by the field depicted in Figure \ref{fig:dsrcpghh0}
  ($f_s=-2P_sp^+$) and zero pressure source (constitutive defect)
  ($h_s=0$).}

\plot{recphh0}{width=\textwidth}{Pressure gather from point source at
  $z_s=3000$ m, $x_s=3500$ m, homogeneous model. Mute applied near cable ends.}

\plot{srcvzghh0}{width=\textwidth}{Approximate inversion of the gather
  in Figure \ref{fig:recphh0} suggested by the first equation in
  display \ref{eqn:approxinv}. Scaled version of output $v_z$ field
  obtained by applying transpose of $V^+_{z_s,z_r}$.}

\plot{fwdphh0}{width=\textwidth}{Pressure gather produced from source
  shown in Figure \ref{fig:srcvzghh0}.}

\plot{diffrecphh0}{width=\textwidth}{Difference between gathers
  displayed in Figures \ref{fig:recphh0} and \ref{fig:fwdphh0},
  plotted on same color scale.}

\plot{invvzghh0}{width=\textwidth}{Inverted pressure source (= scaled
  $v_z$) gather obtained by another application of $(V^+_{z_s,z_r})^T$
  to data shown in Figure \ref{fig:fwdphh0}. Compare Figure
  \ref{fig:srcvzghh0}.}
